Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.

### 2024-03-05 article.tex
Address language issues
- Page 1, Line 49: "Python-centric approach" -> "Our Python-centric approach"?
- Page 2, Line 47: "could be a demanding"
- Page 3, Line 39: "Poincare allows defining" -> "Poincare allows one to define..."
- Page 5, Line 48: "Oscilator" -> "Oscillator"
- Page 5, Line 50: "such that allows to easily define of"
- Page 7, Line 24: "on a factor" -> "by a factor"

\begin{document}
\maketitle
\begin{abstract}
Chemical Reaction Networks (CRNs) play a pivotal role in diverse fields
such as systems biology, biochemistry, chemical engineering, and
epidemiology. High-level modelling of CRNs enables various simulation
approaches, including deterministic and stochastic methods. However,
existing Python tools for CRN modelling typically wrap external C/C++
libraries for modelling and simulation, limiting their extensibility and
integration with the broader Python ecosystem. In response, we developed
Poincaré and SimBio, two novel Python packages for the definition and
simulation of dynamical systems and CRNs. Poincaré serves as a
foundation for dynamical system modelling, while SimBio extends this
functionality to CRNs, including support for the Systems Biology Markup
Language (SBML). Poincaré and SimBio are developed as pure Python
packages enabling users to easily extend their simulation capabilities
by writing new or leveraging other Python packages. Moreover, this does
not compromise the performance, as code can be Just-In-Time compiled
with Numba. Our benchmark tests using curated models from the BioModels
repository, demonstrate that these tools may provide a potentially
superior performance advantage, compared to other existing tools.
Additionally, to ensure a user-friendly experience, our packages use
standard typed modern Python syntax that provides a seamless integration
with Integrated Development Environments (IDEs). Python-centric approach
significantly enhances code analysis, error detection, and refactoring
capabilities, positioning Poincaré and SimBio as valuable tools for the
modelling community.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi

@@

Another challenge is the way models are written. Many tools use a Domain
Specific Language (DSL) for coding, and support the System Biology
Markup Language (SBML) as an exchange format (Hucka et al., n.d.), as
direct SBML coding is impractical. A DSL allows reuse of code in
different programming environments, but it will not be recognized by
default in Integrated Development Environments (IDEs). Therefore DSLs
cannot provide the development experience such as syntax highlighting,
code completion, refactoring, and static analysis, unless such support
is specifically developed. Tellurium uses a Domain Specific Language
(DSL) called Antimony (Smith et al. 2009), which can be translated to
SBML. An extension for Visual Studio Code was developed, but its
maintenance could be demanding task for the systems biology community.
PySB, using Python's dynamic nature, created a DSL within Python. By
default, it uses global state to create species and parameters, without
assigning them to Python variables or adding them explicitly to the
model, but this approach is not fully compatible with IDEs, affecting
the development experience.

To overcome these limitations, we developed poincaré and SimBio,
open-source Python packages for defining and simulating systems.
Poincaré allows defining differential equation systems, while SimBio
builds on it for defining reaction networks. They are focused on
providing an ergonomic experience to end-users by integrating well with
IDEs and static analysis tools through the use of standard modern Python
syntax. Since they are coded in Python, every part from model definition
to simulation can be extended from Python syntax. Being the first-ever
pure Python packages for systems modelling, they offer extensive
extensibility, from simple tasks like reusing integrators defined in
other packages, to complex ones like altering the compilation process to
leverage some structure in the equations. For example, using a for-loop
in the compiled equations could improve the runtime performance if there
is some repetitive structure in the system, as happens in spatial
modelling. The models built using these packages can be introspected to
create other representations, such as graphs connecting species and/or
reactions, or tables with parameters or equations. Furthermore, they
have a modular architecture with a clear separation of concerns, making
it easier to maintain or to contribute new code, which is beneficial for
developers and maintainers. We showcased the reliability of these tools
by benchmarking them against the simulation results from other tools. We
also highlighted the substantial performance improvements our tools
offer, as this is crucial for construction and simulation of models of
whole cells and organisms, which necessitate the simulation of
significantly large-scale models.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The variable name to which a component is assigned can be
  automatically saved in the component for introspection (i.e.,
  \texttt{Oscilator.x.name\ ==\ "x"}),
\item
  It provides a namespace such that allows to easily define of multiple
  independent models in the same script,
\item
  It allows IDEs to provide autocomplete and refactoring capabilities
  (\texttt{Oscillator.\textless{}TAB\textgreater{}} shows \texttt{x},
  \texttt{v} and \texttt{eq}),
\item
  It allows creation of instances which can be composed into a bigger
  model (Figure~\ref{fig-composition}).
\end{enumerate}

@@

Switching backends to \texttt{"numba"} can improve model runtime on a
factor up to x30 or more for big or long running models, however incurs
in a compile time penalty for the first run which must be taken into
account.

>>>>

\begin{document}
\maketitle
\begin{abstract}
Chemical Reaction Networks (CRNs) play a pivotal role in diverse fields
such as systems biology, biochemistry, chemical engineering, and
epidemiology. High-level modelling of CRNs enables various simulation
approaches, including deterministic and stochastic methods. However,
existing Python tools for CRN modelling typically wrap external C/C++
libraries for modelling and simulation, limiting their extensibility and
integration with the broader Python ecosystem. In response, we developed
Poincaré and SimBio, two novel Python packages for the definition and
simulation of dynamical systems and CRNs. Poincaré serves as a
foundation for dynamical system modelling, while SimBio extends this
functionality to CRNs, including support for the Systems Biology Markup
Language (SBML). Poincaré and SimBio are developed as pure Python
packages enabling users to easily extend their simulation capabilities
by writing new or leveraging other Python packages. Moreover, this does
not compromise the performance, as code can be Just-In-Time compiled
with Numba. Our benchmark tests using curated models from the BioModels
repository, demonstrate that these tools may provide a potentially
superior performance advantage, compared to other existing tools.
Additionally, to ensure a user-friendly experience, our packages use
standard typed modern Python syntax that provides a seamless integration
with Integrated Development Environments (IDEs). Our Python-centric approach
significantly enhances code analysis, error detection, and refactoring
capabilities, positioning Poincaré and SimBio as valuable tools for the
modelling community.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi

@@

Another challenge is the way models are written. Many tools use a Domain
Specific Language (DSL) for coding, and support the System Biology
Markup Language (SBML) as an exchange format (Hucka et al., n.d.), as
direct SBML coding is impractical. A DSL allows reuse of code in
different programming environments, but it will not be recognized by
default in Integrated Development Environments (IDEs). Therefore DSLs
cannot provide the development experience such as syntax highlighting,
code completion, refactoring, and static analysis, unless such support
is specifically developed. Tellurium uses a Domain Specific Language
(DSL) called Antimony (Smith et al. 2009), which can be translated to
SBML. An extension for Visual Studio Code was developed, but its
maintenance could be a demanding task for the systems biology community.
PySB, using Python's dynamic nature, created a DSL within Python. By
default, it uses global state to create species and parameters, without
assigning them to Python variables or adding them explicitly to the
model, but this approach is not fully compatible with IDEs, affecting
the development experience.

To overcome these limitations, we developed poincaré and SimBio,
open-source Python packages for defining and simulating systems.
Poincaré allows one to define differential equation systems, while SimBio
builds on it for defining reaction networks. They are focused on
providing an ergonomic experience to end-users by integrating well with
IDEs and static analysis tools through the use of standard modern Python
syntax. Since they are coded in Python, every part from model definition
to simulation can be extended from Python syntax. Being the first-ever
pure Python packages for systems modelling, they offer extensive
extensibility, from simple tasks like reusing integrators defined in
other packages, to complex ones like altering the compilation process to
leverage some structure in the equations. For example, using a for-loop
in the compiled equations could improve the runtime performance if there
is some repetitive structure in the system, as happens in spatial
modelling. The models built using these packages can be introspected to
create other representations, such as graphs connecting species and/or
reactions, or tables with parameters or equations. Furthermore, they
have a modular architecture with a clear separation of concerns, making
it easier to maintain or to contribute new code, which is beneficial for
developers and maintainers. We showcased the reliability of these tools
by benchmarking them against the simulation results from other tools. We
also highlighted the substantial performance improvements our tools
offer, as this is crucial for construction and simulation of models of
whole cells and organisms, which necessitate the simulation of
significantly large-scale models.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The variable name to which a component is assigned can be
  automatically saved in the component for introspection (i.e.,
  \texttt{Oscillator.x.name\ ==\ "x"}),
\item
  It provides a namespace that allows to easily define multiple
  independent models in the same script,
\item
  It allows IDEs to provide autocomplete and refactoring capabilities
  (\texttt{Oscillator.\textless{}TAB\textgreater{}} shows \texttt{x},
  \texttt{v} and \texttt{eq}),
\item
  It allows creation of instances which can be composed into a bigger
  model (Figure~\ref{fig-composition}).
\end{enumerate}

@@

Switching backends to \texttt{"numba"} can improve model runtime by a
factor up to x30 or more for big or long running models, however incurs
in a compile time penalty for the first run which must be taken into
account.
